---
title: "Writing Robust R Functions"
description: |
  Some designs to validate function arguments.
author:
  - name: Vishal Katti
    url: {}
date: 2022-01-18
preview: https://i.imgur.com/hbjbLMN.gif
categories:
  - Rstats
  - functions
output:
  distill::distill_article:
    self_contained: false
    toc: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comments = '')
knitr::knit_hooks$set(error = function(x, options) {
  paste0("<pre style=\"color: red;\"><code>", x, "</code></pre>")
})
knitr::knit_hooks$set(warning = function(x, options) {
  paste0("<pre style=\"color: orange;\"><code>", x, "</code></pre>")
})
```

## Introduction

Functions in R ( or any other programming language in general) allow us
to encapsulate some lines of code that we want to run again and again.
Functions are the natural outcome of the **DRY** (Don't Repeat
Yourself!) principle. Functions group together a couple of lines of
consistent logic making our code modular and consequently, easy to
manage. However, when we write functions, we need to ensure that they
behave exactly as we want them to and are able to handle whatever we
throw at them. By whatever, I mean any and all kinds of inputs. The idea
of creating unbreakable code is idealistic. I say this since creating
robust functions requires additional code to handle the unwanted inputs
and most useRs write functions during some one-time analysis. Hence we
need to be pragmatic about how much time and effort we spend trying to
make our functions robust. Maybe, we need our functions to be just
robust enough! All I am saying is, if you are creating functions that
will be used by you and only you i.e. if you have absolute control over
what inputs would be provided to your functions, then you can forego
certain checks and the functions need not be unbreakable. But, if you
intend to write functions that will be used by a larger audience, you
need to ensure that such functions are able to handle all kinds of
innocent and malicious intents.

## What do we mean by Robust Functions?

You must be familiar with the *Garbage-In-Garbage-Out* philosophy of
Software engineering. We can think of it in terms of functions, that,
given garbage or bad input, you get garbage or bad output. For a
function to be robust, it must behave in a consistent manner for known
and correct inputs, however, more importantly, it mustn't give us
garbage for bad inputs. Rather, it must provide useful output (as
messages or instructions) which can be further used to inform the
end-user about possible problems in the inputs to drive proper usage.
The useful output/s in case of bad inputs would ideally be a combination
of clean early exit and easy-to-understand error messages. So we shall
try to implement *Garbage-In-Useful-Info-Out* by looking at some ways we
can build well-behaved and reliable functions.

Input values passed to a function are more popularly known as arguments
or parameters. A robust function must validate the function arguments
before proceeding to implement the function logic. If this is not done,
then the bad arguments will cause some errors in the logic and display
error messages that the end-user may not be familiar with. Worst-case
scenario is when the function doesn't encounter any errors and just
gives bad results!! Surely, we do not want this unpredictable behavior.

![Enough Talk, Let's Fight! - Kungfu Panda
\@imgflip.com](https://i.imgflip.com/28bxz3.jpg){width="60%"}

## Our sweet, innocent and naive Function

Consider the following function `make_date` that takes 3 numeric inputs
`yyyy`, `mm` and `dd` and returns a single 'Date\` object.

```{r}
make_date <-  function(yyyy, mm, dd) {
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

my_date <- make_date(yyyy = 2022, mm = 1, dd = 31)
my_date
class(my_date)
```

We will use `make_date` to demonstrate a couple of scenarios where this
function can fail and the methods to safeguard against such scenarios.

## Scenario 1: Missing Arguments

The most basic check we should perform before running the function logic
is to confirm if all the required arguments are available. Think about
how your function should behave if one of the arguments, suppose `mm` is
missing.

```{r error=TRUE}
make_date(yyyy = 2022, dd = 31)
```

Note that the error message shown to the user, is triggered, not from
our function `make_date` but from the internal `paste` function. We do
not have any control over what error messages are shown when errors
occur. In this case, we know specifically that this error is due to a
missing argument.

There are two ways to handle missing arguments:

### 1.1 Early Exit

If a certain required argument is missing, we can stop the execution of
the function and show informative error message about which argument is
missing. Your friends here are the `missing` and `stop` functions. The
`missing` function checks if the given argument is missing or is set to
NULL and returns TRUE, else it returns FALSE. The `stop` function stops
the execution and displays the custom error message we provide. Using
these functions inside an `if` condition will let us check for missing
arguments. Let us modify our naive function to stop early when required
arguments are missing.

```{r error=TRUE}
make_date <-  function(yyyy, mm, dd) {
  
  # check missing arguments
  if (missing(yyyy)) stop("argument `yyyy` is required.")
  if (missing(mm))   stop("argument `mm` is required.")
  if (missing(dd))   stop("argument `dd` is required.")
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

# Calling the function without `mm` argument
make_date(yyyy = 2022, dd = 31)
```

Note that here, we add three `if-missing-stop` statements, one for each
required argument. We must do this if we want to display specific error
messages for each argument. There is another way to do the same but we
will look at it later. If we want to display a single error message, we
can do so by clubbing the `missing` functions inside an `any` which will
return TRUE if any one of the arguments is missing. However, providing
clear error messages becomes challenging in this method.

```{r error=TRUE}
dummy_fun <- function(a, b, c) { 
  if(any(missing(a), missing(b), missing(c))) {
    stop("One or more required arguments missing.")
  }
  # Do something...
}
dummy_fun(a = 1)
```

### 1.2 Sensible defaults with warnings

In some cases, we may need the function to use some sensible default
value for the required arguments and continue execution. Here, we
display a warning message instead of an error message. This is required
when the argument value is either considered to be obvious or the
argument is not necessarily the most important one and is used only in
extreme customization. **Providing default values to arguments makes
them optional arguments**. An example of default argument values can be
seen in the `paste` function we have used earlier. The default value of
the separator argument `sep` is a single whitespace character.

```{r}
args(paste)
```

Similarly, we can provide some sensible defaults for the `make_date`
function. Let's modify the function further to provide defaults for the
`mm` and `dd` arguments only.

```{r warning=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

# Calling the function without `mm` and `dd` arguments
make_date(yyyy = 2022) # here, only `yyyy` is the required argument.

```

There are a few concerns about using warnings instead of error messages.
Some are listed here in this article from RBloggers [A Warning About
`warning`](https://www.r-bloggers.com/2012/05/a-warning-about-warning/).

## Scenario 2: Invalid Argument Data Type

We have defined `make_date` to accept 3 numeric arguments i.e. all 3
must be numbers. What would happen if someone tried to call `make_date`
with character, factor or boolean inputs?

```{r}
make_date(yyyy = "2022", mm = "5", dd = "20") # works!! why?
```

In this case, the function works because when the arguments are combined
into a single string using `paste` , it matches the `format` argument of
the `as.Date` function in the main logic of `make_date` which is
`as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")`

```{r}
make_date(yyyy = "2022", mm = "May", dd = "1") # works but shows NA !!!
```

In this case, all the arguments pass the checks but the output is `NA`
since we pass `2022-May-1` to `as.Date` which doesn't match the
`format = '%Y-%m-%d'`.

How do we check if the values provided to the arguments are numbers or
number-like? If the values are numbers, we let them pass. But if they
are non-numeric, we must check if they can be converted to numbers i.e.
we must check if they are number-like. By number-like, I mean, will
coercing the value using `as.numeric` give us a numeric value or `NA` ?
You guessed it right, we will pass the values through `as.numeric` and
check if the output is `NA` or not.

What are the various data types in R that are not `numeric` but can look
like numbers? We have `character`, `factor` and `boolean` data types
which can behave like numbers sometimes. Let's see a few scenarios.

### Character arguments

```{r}
Year <- c("2022", "TwentyTwo")
Year_num <- as.numeric(Year) # this should show a warning about NAs introduced by coercion
Year_num # must show the number 2022 without quotes and one NA
```

As you can see in above example, when passed through `as.numeric`, the
value "2022" gets converted to the number 2022 but the value "TwentyTwo"
does not. Hence we can say "2022" is number-like but "TwentyTwo" is not.

### Factor arguments

```{r}
Year <- factor(c("2022","2021","TwentyTwo"))
as.numeric(Year)

YearX <- factor(c("2022", "X"))
as.numeric(YearX)

YearY <- factor(2022)
as.numeric(YearY)
```

As you can see from above examples, `factor` values do get converted to
numeric but do not give the right results. So we can safely say that
factors are not number-like.

I will ignore `boolean` data types hoping that useRs are bright enough
to not use Booleans while creating a Date!

From the above examples, we can conclude that `numeric` values and
number-like `character` values are the only valid data types that should
be allowed. Modifying our `make_date` function to include data type
checks.

```{r error=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  # Check data types
  if (!is.numeric(yyyy) & !is.character(yyyy)) {
    stop("argument `yyyy` must be numeric")
  } else if (is.character(yyyy) & is.na(as.numeric(yyyy))) {
    stop("argument `yyyy` must be numeric")
  }
  if (!is.numeric(mm) & !is.character(mm)) {
    stop("argument `mm` must be numeric")
  } else if (is.character(mm) & is.na(as.numeric(mm))) {
    stop("argument `mm` must be numeric")
  }
  if (!is.numeric(dd) & !is.character(dd)) {
    stop("argument `dd` must be numeric")
  } else if (is.character(dd) & is.na(as.numeric(dd))) {
    stop("argument `dd` must be numeric")
  }
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

# Calling the function with new datatype checks
make_date(yyyy = "2022", mm = "May", dd = "1")
make_date(yyyy = "2022", mm = factor("5"), dd = "1")
```

Notice that the datatype check is lengthy and similar for all 3
arguments. We can apply **DRY** principle again and encapsulate that
code into a small function `is_numberlike` which will return `TRUE` or
`FALSE` . Note that `is_numberlike` has no checks because it is an
internal function.

```{r}
# This function check if value is number or number-like.
is_numberlike <- function(x){
  if (!is.numeric(x) & !is.character(x)) {
    return(FALSE) # Early Exit 1 if value is neither numeric nor character
  } else if (is.character(x) & is.na(as.numeric(x))) {
    return(FALSE) # Early Exit 2 if character value is not number-like.
  }
  return(TRUE)
}
```

Thus our `make_date` function with data types check will look as below.

```{r error=TRUE, warning=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  # Check data types
  if (!is_numberlike(yyyy)) stop("argument `yyyy` must be numeric")
  if (!is_numberlike(mm))   stop("argument `mm` must be numeric")
  if (!is_numberlike(dd))   stop("argument `dd` must be numeric")
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

# Calling the function with new datatype checks
make_date(yyyy = "TwentyTwo", mm = "5", dd = 1)
make_date(yyyy = "2022", mm = factor("5"), dd = "1")
make_date(yyyy = 2022, mm = 5, dd = "one")

```

One of the most interesting features of R is vectorization! Due to this
feature, our function `make_date` behaves in interesting ways. In some
cases, it is desirable and sometimes it is not.

```{r}
make_date(yyyy = 2022, mm = 1:12, dd = "1")
```

Note the above warnings. These warnings appear because the `if`
statement checks if the condition provided results in a single `TRUE` or
`FALSE` value. However, the output of the check `is.na(as.numeric(mm))`
is a boolean vector of length 12. But `if` needs only 1 `TRUE` or
`FALSE`.

The output contains 12 date values since `paste` is vectorised, it
recycles the values for `yyyy` and `dd` to give us 12 dates!

```{r}
mm <- 1:12
paste("Month", mm)
```

What do we do if we want `make_date` to return just one date?

## Scenario 3: Incorrect Argument Size

To ensure `make_date` gives you just one date, we must ensure that the
arguments have just value and is not a vector of multiple values i.e.
`length(arg)==1`. Let's further add a few checks for the data size of
the arguments and rearrange the checks.

```{r error=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  # Check argument lengths
  if (length(yyyy)!=1) stop(paste0("Length of argument `yyyy` is ", length(yyyy),". Must be only 1."))
  if (length(mm)!=1)   stop(paste0("Length of argument `mm` is ", length(mm),". Must be only 1."))
  if (length(dd)!=1)   stop(paste0("Length of argument `dd` is ", length(dd),". Must be only 1."))
  
  # Check data types
  if (!is_numberlike(yyyy)) stop("argument `yyyy` must be numeric")
  if (!is_numberlike(mm))   stop("argument `mm` must be numeric")
  if (!is_numberlike(dd))   stop("argument `dd` must be numeric")
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

# Calling function with new data size checks
make_date(yyyy = 2022, mm = 1:12, dd = "01")
make_date(yyyy = c("2021","2022"), mm = "1", dd = 1)
make_date(yyyy = 2022, mm = 1, dd = c("1","2"))
```

## A little detour...

So far we checked for missing arguments, arguments with bad data types
and arguments with incorrect sizes. We've used the `stop` function along
with `if` to check for all failure conditions and show appropriate error
messages. When we use `stop`, we must specify all the failure conditions
and the number of specific error messages goes up as number of arguments
increases.

In case of our `make_date`, if an argument is not missing, it must be a
number-like value of length 1. To reduce the number of error messages,
we can combine the error messages for data type and length. for eg, the
error message could be *argument `yyyy` must be a number-like value of
length 1.*

Wouldn't it be easier if we just specify what is the success condition
aka the "happy path", and show error for all other conditions? To do
this, we can use the `stopifnot` function that let's us specify all the
happy paths. See example below.

```{r error=TRUE}
dummy_sum <- function(a, b, c){
  
  # check missing
  stopifnot(!missing(a) & !missing(b) & !missing(c))
  
  # check argument values
  stopifnot(!is.na(a) & is.numeric(a) & length(a)==1,
            !is.na(b) & is.numeric(b) & length(b)==1,
            !is.na(c) & is.numeric(c) & length(c)==1
            )
  sum(a, b, c)
}

dummy_sum(b = 2, c = 3) # a is missing
dummy_sum(a = NA_integer_, b = 2, c = 3) # a has NA value
dummy_sum(a = 1, b = "2", c = 3) # b has non-numeric value
dummy_sum(a = 1, b = 2, c = 5:7)  # c has length != 1

```

Note the error messages above. They are not so user-friendly. Luckily,
we can specify error messages in `stopifnot` by providing the error
messages as the names of the "happy path" conditions.

```{r error=TRUE}
dummy_sum <- function(a, b, c){
  
  # check missing
  stopifnot("one or more required arguments missing" = !missing(a) & !missing(b) & !missing(c))
  
  # check argument values
  stopifnot("argument `a` must not be NA, must be a number of length 1" = !is.na(a) & is.numeric(a) & length(a)==1,
            "argument `b` must not be NA, must be a number of length 1" = !is.na(b) & is.numeric(b) & length(b)==1,
            "argument `c` must not be NA, must be a number of length 1" = !is.na(c) & is.numeric(c) & length(c)==1
            )
  sum(a, b, c)
}

dummy_sum(b = 2, c = 3) # a is missing
dummy_sum(a = NA_integer_, b = 2, c = 3) # a has NA value
dummy_sum(a = 1, b = "2", c = 3) # b has non-numeric value
dummy_sum(a = 1, b = 2, c = 5:7)  # c has length != 1

```

Using `stopifnot` in our `make_date` function to combine the datatype
and length checks, we get...

```{r warning=TRUE, error=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  
  # Check argument types and length
  stopifnot(
    "argument `yyyy` must be numeric with length 1" = is_numberlike(yyyy) & length(yyyy)==1,
    "argument `mm` must be numeric with length 1"   = is_numberlike(mm)   & length(mm)==1,
    "argument `dd` must be numeric with length 1"   = is_numberlike(dd)   & length(dd)==1
  )
  
  # main logic
  as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
}

make_date() # no arguments provided
make_date(yyyy = 2022, mm = 1:12, dd = 31) # Length mm not equal to 1
make_date(yyyy = 2022, mm = "Jan", dd = 31) # mm is not number-like
make_date(yyyy = 2022, dd = 31) # argument mm is missing but should work using default value


```

## Scenario 4: Values of Arguments that result in invalid outputs

Finally, what do we do when the arguments provided will definitely give
us bad results despite passing all checks? In our case, `make_date`
creates a date but if we give it values that will result in an invalid
date, it will give us invalid results (remember
*Garbage-In-Garbage-Out*?).

```{r}
make_date(yyyy = 2022, mm = 13, dd = 1) # is there a 13th month?
```

We get `NA` because `as.Date` returns NA for invalid inputs with no
error messages or warnings! We can check the output and provide a
generic error message.

```{r warning=TRUE, error=TRUE}
make_date <-  function(yyyy, mm = 1, dd = 1) {
  # check missing arguments
  if (missing(yyyy))  stop("argument `yyyy` is required.") 
  if (missing(mm)) warning("argument `mm` is missing. Using default value mm = 1 instead") 
  if (missing(dd)) warning("argument `dd` is missing. Using default value dd = 1 instead")
  
  
  # Check argument types and length
  stopifnot(
    "argument `yyyy` must be numeric with length 1" = is_numberlike(yyyy) & length(yyyy)==1,
    "argument `mm` must be numeric with length 1"   = is_numberlike(mm)   & length(mm)==1,
    "argument `dd` must be numeric with length 1"   = is_numberlike(dd)   & length(dd)==1
  )
  
  # main logic
  out <- as.Date(paste(yyyy, mm, dd, sep = "-"), format = "%Y-%m-%d")
  if (is.na(out)) {
    stop("Invalid values provided. Please check your inputs.")
  }
  return(out)
}

make_date(yyyy = 2022, mm = 13, dd = 1) # is there a 13th month?
make_date(yyyy = 2022, mm = 2, dd = 31) # are there 31 days in February?
```

Do you think our function `make_date` is robust enough?

![As robust as Superman! Source:
Imgur](https://i.imgur.com/hbjbLMN.gif)

## Conclusion

Making functions robust requires some prior thought about its intended
use and audience. Based on this, we can decide what checks to implement,
what to skip, whether to stop execution using error messages or to use
default values with warnings. Checking for "happy paths" is simpler
compared to checking each and every bad input and providing specific
error messages. Too many different error messages for the same argument
could become a source of frustration of the end user, so consider
combining some checks and their error messages to be informative and
precise. Robustness, like everything else, in moderation, is good and
getting it "just right" takes time and dedicated effort. Happy Coding!

## Citations & References

-   [Techniques for writing robust R programs -
    LexJansen](https://www.lexjansen.com/phuse/2014/ts/TS03.pdf)
-   [R Programming for Data
    Science](https://bookdown.org/rdpeng/rprogdatascience/functions.html)
-   [A Warning About
    warning](https://www.r-bloggers.com/2012/05/a-warning-about-warning/)
